A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-11-15 analysis/paper/paper.Rmd
Revised manuscript

Keywords: `r rmarkdown::metadata$keywords`
```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  fig.align = 'center',
  dpi = 300
)

@@

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has caused most areas of Norway to have been subjected to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @mörner1979; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is faster towards the centre of the ice sheet relative to the distal aspects. Thus, there is differential glacio-isostatic impact to a site’s location depending on it's relation to the ice sheet’s centre of mass, leading some areas on the outer coast to have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time spans, creating a mix of settlement phases that are difficult to disentangle [e.g. @hagen1963; @reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, submerge them [@bjerck2008; @glorstad2020]. Transgression can therefore lead to a hiatus in the archaeological record for certain sub-phases in the impacted areas despite the fact that there were likely coastal settlements during the inferred hiatuses. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement pattern over a long and continuous time span.

```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c. 17--8 thousand years (kyr) BCE. The study area defined later in the text is marked with a red outline (deglaciation data from Hughes et al. 2016, but see also Romundset et al. 2019 in relation to the study area).", out.width = "525px", fig.align = "center", regfloat = TRUE}

@@

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, it has been characterised as a relative dating method for sites located within a constrained geographical area, or it has been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999]. The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @glorstad2002; @glorstad2003; @glorstad2004; @jaksland2001; @jaksland2012; @jaksland2012a; @jaksland2014a; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100-, 200- or 500-year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020]. In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method---what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963; @johansen1997; @mikkelsen1975a, 100].

@@

One of the more extensive evaluations of the relationship between archaeological radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.3% of the dates [@solheim2020, 48]. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these dates are entirely erroneous can be difficult to evaluate [e.g. @persson2008; @schulke2020]. However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas as evidenced by artefact inventories or multiple site features. Not remnants of stays as ephemeral to only be discernible by isolated features or dubious ^14^C-dates. The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent upon, and follow the discretion of the original excavation reports.

@@

The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2018, https://hoydedata.no]. The 10m resolution DTM was used rather than the higher-resolution 1m version, both because this resulted in considerably less processing time and because the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances. The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018]. All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in a version-controlled repository at <https://osf.io/7f9su/>, organised as a research compendium following Marwick et al. [-@marwick2018; @Marwick2017].

@@

Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with temporal uncertainty. The first task was therefore to ascribe a likely date and associated degree of uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the sites all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if their date ranges intersected at 99.7% probability, meaning these were effectively taken to be associated with the same event, here termed settlement or site phase. In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed using the Sum function. Multiple phases at a single site were treated as independent of each other.

@@

In the case where the sea polygons intersect the site polygon, all distance measures were set to zero. In the case that the sea polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located below sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values were therefore retained with the exception of the sites of Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).

This process was repeated 1000 times for each phase for each site (Figure \@ref(fig:example-3)). The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5 [cf. @crema2010, 1125]. This evaluation is presented in the supplementary material. Hovland 5 was chosen for this assessment as it has an imprecise age and is located in area of quite complex topography.

@@

```{r example-2, echo = FALSE, fig.cap = "Example of a single simulation run on the site Hegna vest 1. A) The simulation starts by drawing a single year, weighted by the the posterior probability distribution.  B) This then corresponds to an elevation range on the interpolated displacement curve. A single elevation is drawn uniformly from this range using 5cm intervals. C) The sea-level is then adjusted on the DTM to this elevation and the various distance measures are found. D) The numerical result of the simulation run.", out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/example_sim_map.png"))
```

@@

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. @berg-hansen2022[644] made a similar observation in their assessment of the correspondence between shoreline displacement and radiocarbon dates from these sites. The sites are recently excavated, well-dated and situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022]. While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent  
an upper constraining limit for the sea-level, as they would not have been in use when located under water. It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.

@@

```{r results2, echo = FALSE, fig.cap = paste0("Histograms showing the simulated distance from the shoreline using radiocarbon dates corresponding to the site inventories. Negative values have been set to zero. A) Simulated results older than 2500 BCE (", nsitesdist25 , " sites and ", nphasesdist25 ," site phases) and B) simulated results older than 4000 BCE (", nsitesdist4 , " sites and ", nphasesdist4 ," site phases). Note that the cut-off is done based on the calendar year associated with each distance value. Consequently, sites and site phases are only completely excluded if the entire posterior probability of the radiocarbon dates falls later than the cut-off. Furthermore, the superimposed gamma distributions have been fit when adding a constant of 0.001 to the distance values and have been cut off on the y-axis for visualisaton. The gamma distribution in A forms the basis for the analysis to follow, but a version has also been fit to the vertical distances in B to further illustrate the difference between the distributions."), out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/results2.png"))
```

@@

An example of an implementation of the outlined approach is given in Figure \@ref(fig:exampledat), where $\alpha$ = `r round(gammafit$estimate[1], 3)` and $\sigma$ = `r round(gammafit$estimate[2], 3)`. These are the parameters for the gamma distribution identified when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A) and are the parameters used in all applications of the proposed method that follow below. For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, which, following from the adjustment of the values for fitting the compared models, starts from 0.001m. The gamma distribution is stepped through in it's cumulative form, where the probability from the previous 0.001m step is subtracted from the probability at the current step. This probability is then divided equally across the individual calendar years in the range between the lower and the upper limit of the displacement curve at the current 0.001m step. The probability mass function that is the resulting shoreline date is the sum of performing this procedure on all possible 0.001m values of $D$, which, in practice, is down to and including $E - D = 0.001$ or when 99.999% of the gamma distribution has been stepped through.

@@

The small discrepancies between the achieved results mainly follow from the fact that a slightly updated version of the local displacement curve is applied here [cf. @sørensen2014; @sorensen2023]. Jaksland's dates are given a flat 200- and 50-year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while his argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they do not overlap. This information is not integrated in the approach outlined here, but it could justify further reducing the uncertainty ranges. Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates.

Above it was suggested that additional temporal data could be combined with the method to improve its precision. Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help to increase the precision of the method if this can be used to exclude certain RSLs as unlikely for when a site was in use. One potential way to do this could be through the analysis of phosphate concentrations in soils, which has the potential to offer insights on the likely position of the shoreline when a site was in use [@ilves2011]. This has been done in the Baltic sea region [e.g. @broadbent1979; @sundström2006; @ilves2011], but has yet to provide reliable results in Norway [e.g. @melvold2014a; @viken2018b]. The identification of other physical traces of shore formation processes and the deposition of beach sediments in relation to archaeological material also holds similar potential [e.g. @bondevik2019]. Finally, another approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline. If such methods are developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use. Such approaches would make less of an impact for the Brunlanes sites, where the 95% HDRs are already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf. Figure \@ref(fig:redate)).

```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating of the Brunlanes sites using site altitudes provided by Jaksland (2014:tab.4). A) The result of applying the approach to shoreline dating outlined above. The shoreline date in grey is underlined with the 95\\% HDR in black. Dates provided by Jaksland (2014) are plotted in red. The box indicates a 50 year uncertainty range which in combination with the red line extends 200 years. B) Map showing the centroids of the Pauler sites and Sky 1. The sea-level has been simulated using the probability distribution associated with the shoreline date for Pauler 1 (see also map in Jaksland 2014:fig.12a). Pauler 1 is the red point.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/brunlanes.png"))
```

@@

As the nature of the relationship between sites and sea is likely to vary temporally and geographically [e.g. @nyland2020], the proposed implementation and parametrisation of shoreline dating can not be expected to be directly applicable elsewhere. When this is combined with the fact that the rate of RSL-change also varies geographically and temporally [e.g. @svendsen1987], this means that the accuracy and precision of the method will also vary. However, the methodological framework used to evaluate the relationship between sites and sea is readily extendible to other regions of northern Scandinavia where reliable data on shoreline displacement is available, thus making such extensions feasible. Furthermore, the simulation approach used to integrate multiple sources of spatio-temporal uncertainty was used here to inform the question of the distance between sites and the shoreline. However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent to this reconstruction.

>>>>

Keywords: `r rmarkdown::metadata$keywords`
```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.hide = TRUE,
  fig.path = "../figures/",
  fig.align = 'center',
  dpi = 300
)

@@

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has caused most areas of Norway to have been subjected to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @mörner1979; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is faster towards the centre of the ice sheet relative to the distal aspects. Thus, there is differential glacio-isostatic impact to a site’s location depending on its relation to the ice sheet’s centre of mass, leading some areas on the outer coast to have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time spans, creating a mix of settlement phases that are difficult to disentangle [e.g. @hagen1963; @reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, submerge them [@bjerck2008; @glorstad2020]. Transgression can therefore lead to a hiatus in the archaeological record for certain sub-phases in the impacted areas despite the fact that there were likely coastal settlements during the inferred hiatuses. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement pattern over a long and continuous time span.

```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000-year intervals from c. 17--8 thousand years (kyr) BCE. The study area defined later in the text is marked with a red outline (deglaciation data from Hughes et al. 2016, but see also Romundset et al. 2019 in relation to the study area).", out.width = "525px", fig.align = "center", regfloat = TRUE}

@@

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, it has been characterised as a relative dating method for sites located within a constrained geographical area, or it has been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999]. The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @glorstad2002; @glorstad2003; @glorstad2004; @jaksland2001; @jaksland2012; @jaksland2012a; @jaksland2014a; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b; @solheim2013 and below]. Recently the method has also been used independently to date a larger number of sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100-, 200- or 500-year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020]. In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method---what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963; @johansen1997; @mikkelsen1975a, 100].

@@

One of the more extensive evaluations of the relationship between archaeological radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.3% of the dates [@solheim2020, 48]. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these dates are entirely erroneous can be difficult to evaluate [e.g. @persson2008; @schulke2020]. However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas as evidenced by artefact inventories or multiple site features. Not remnants of stays as ephemeral to only be discernible by isolated features or dubious ^14^C-dates. The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent upon and follow the discretion of the original excavation reports.

@@

The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2018, https://hoydedata.no]. The 10m resolution DTM was used rather than the higher-resolution 1m version, both because this resulted in considerably less processing time and because the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances. The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018]. All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in a version-controlled repository at <https://osf.io/7f9su/>, organised as a research compendium following Marwick [@marwick2018; -@Marwick2017].

@@

Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with temporal uncertainty. The first task was therefore to ascribe a likely date and associated degree of uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the sites all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if their date ranges intersected at 99.7% probability, meaning these were effectively taken to be associated with the same occupation event, here termed settlement or site phase. In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed using the Sum function. Multiple phases at a single site were treated as independent of each other.

@@

In the case where the sea polygons intersect the site polygon, all distance measures were set to zero. In the case that the sea polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located below sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values were therefore retained except of for the sites Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).

This process was repeated 1000 times for each phase for each site (Figure \@ref(fig:example-3)). The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5 [cf. @crema2010, 1125]. This evaluation is presented in the supplementary material. Hovland 5 was chosen for this assessment as it has an imprecise age and is located in area of quite complex topography [@mansrud2013].

@@

```{r example-2, echo = FALSE, fig.cap = "Example of a single simulation run on the site Hegna vest 1. A) The simulation starts by drawing a single year, weighted by the posterior probability distribution.  B) This then corresponds to an elevation range on the interpolated displacement curve. A single elevation is drawn uniformly from this range using 5cm intervals. C) The sea-level is then adjusted on the DTM to this elevation and the various distance measures are found. D) The numerical result of the simulation run.", out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/example_sim_map.png"))
```

@@

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. @berg-hansen2022[644] made a similar observation in their assessment of the correspondence between shoreline displacement and radiocarbon dates from these sites. The sites are recently excavated, well-dated and are situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022]. While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent an upper constraining limit for the sea-level, as they would not have been in use when located under water. It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.

@@

```{r results2, echo = FALSE, fig.cap = paste0("Histograms showing the simulated distance from the shoreline using radiocarbon dates corresponding to the site inventories. Negative values have been set to zero. A) Simulated results older than 2500 BCE (", nsitesdist25 , " sites and ", nphasesdist25 ," site phases) and B) simulated results older than 4000 BCE (", nsitesdist4 , " sites and ", nphasesdist4 ," site phases). Note that the cut-off is done based on the calendar year associated with each distance value. Consequently, sites and site phases are only completely excluded if the entire posterior probability of the radiocarbon dates falls later than the cut-off. Furthermore, the superimposed gamma distributions have been fit when adding a constant of 0.001 to the distance values and have been cut off on the y-axis for visualisation. The gamma distribution in A forms the basis for the analysis to follow, but a version has also been fit to the vertical distances in B to further illustrate the difference between the distributions."), out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/results2.png"))
```

@@

An example of an implementation of the outlined approach is given in Figure \@ref(fig:exampledat), where $\alpha$ = `r round(gammafit$estimate[1], 3)` and $\sigma$ = `r round(gammafit$estimate[2], 3)`. These are the parameters for the gamma distribution identified when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A) and are the parameters used in all applications of the proposed method that follow below. For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, which, following from the adjustment of the values for fitting the compared models, starts from 0.001m. The gamma distribution is stepped through in its cumulative form, where the probability from the previous 0.001m step is subtracted from the probability at the current step. This probability is then divided equally across the individual calendar years in the range between the lower and the upper limit of the displacement curve at the current 0.001m step. The probability mass function that is the resulting shoreline date is the sum of performing this procedure on all possible 0.001m values of $D$, which, in practice, is down to and including $E - D = 0.001$ or when 99.999% of the gamma distribution has been stepped through.

@@

The small discrepancies between the achieved results mainly follow from the fact that a slightly updated version of the local displacement curve is applied here [cf. @sørensen2014; @sorensen2023]. Jaksland's dates are given a flat 200- and 50-year uncertainty range starting from what he gives as the earliest possible date. The 200-year uncertainty range is given if the sites were to be considered in isolation, while his argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they do not overlap. This information is not integrated in the approach outlined here, but it could justify further reducing the uncertainty ranges. Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates.

Above it was suggested that additional temporal data could be combined with the method to improve its precision. Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help to increase the precision of the method if this can be used to exclude certain RSLs as unlikely for when a site was in use. One potential way to do this could be through the analysis of phosphate concentrations in soils, which has the potential to offer insights on the likely position of the shoreline when a site was in use [@ilves2011]. This has been done in the Baltic Sea region [e.g. @broadbent1979; @sundström2006; @ilves2011], but has yet to provide reliable results in Norway [e.g. @melvold2014a; @viken2018b]. The identification of other physical traces of shore formation processes and the deposition of beach sediments in relation to archaeological material also holds similar potential [e.g. @bondevik2019]. Finally, another approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline. If such methods are developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use. Such approaches would make less of an impact for the Brunlanes sites, where the 95% HDRs are already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf. Figure \@ref(fig:redate)).

```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating of the Brunlanes sites using site altitudes provided by Jaksland (2014:tab.4). A) The result of applying the approach to shoreline dating outlined above. The shoreline date in grey is underlined with the 95\\% HDR in black. Dates provided by Jaksland (2014) are plotted in red. The box indicates a 50-year uncertainty range which in combination with the red line extends 200 years. B) Map showing the centroids of the Pauler sites and Sky 1. The sea-level has been simulated using the probability distribution associated with the shoreline date for Pauler 1 (see also map in Jaksland 2014:fig.12a). Pauler 1 is the red point.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/brunlanes.png"))
```

@@

As the nature of the relationship between sites and sea is likely to vary temporally and geographically [e.g. @nyland2020], the proposed implementation and parametrisation of shoreline dating cannot be expected to be directly applicable elsewhere. When this is combined with the fact that the rate of RSL-change also varies geographically and temporally [e.g. @svendsen1987], this means that the accuracy and precision of the method will also vary. However, the methodological framework used to evaluate the relationship between sites and sea is readily extendible to other regions of northern Scandinavia where reliable data on shoreline displacement is available, thus making such extensions feasible. Furthermore, the simulation approach used to integrate multiple sources of spatio-temporal uncertainty was used here to inform the question of the distance between sites and the shoreline. However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent to this reconstruction.
